---
title: "Plots and analysis of Pubmed queries"
author: "Jose L. Cadavid"
output: html_notebook
---

This notebook loads tables of pubmed IDs that come from multiple queries performed
in the "run_pubmed_queries" notebook. Most of the chunks can be run fully automatically,
except the last few where user input is needed when looking at article titles and abstracts.

# Load packages

```{r}
library(readxl, quietly = T)
library(scales, quietly = T)
```

# Load data

Define directory where files are contained and run extra functions for plotting and so on.

Note: Query tables last updated September, 2023

```{r}
setwd("E:/Jose Luis/Documents/GitHub/SysBio_TissueEng_Review/Code")
source("extra_functions.R")
# Define directories
query_file <- "../Data/queries.xlsx"
dir_out <- "../Results/"
# Read tables with results from queries
tables <- readRDS(file = paste0(dir_out, "query_tables.rds"))
# Read query terms
my_query_SB <- read_excel(query_file, sheet = "Modeling") %>% filter(!grepl("#", term))
# Get tissue engineering and in vitro models query terms
my_query_TE <- read_excel(query_file, sheet = "TissueEng") %>% filter(!grepl("#", term))
# Get omic and high throughput measurement terms
my_query_Omic <- read_excel(query_file, sheet = "Method") %>% filter(!grepl("#", term))
```

## Intersect PMID categories

Intersect interesting categories to make plots later on or to check interesting IDs manually

```{r}
tables_mixed <- NULL
tables_mixed[["TE_SB_gen"]] <- merge(tables[["TE_gen"]], tables[["SB_gen"]], by = c("year", "PMID"))
tables_mixed[["TE_Omic_gen"]] <- merge(tables[["TE_gen"]], tables[["Omic_gen"]], by = c("year", "PMID"))
# Intersect disease queries
tables_mixed[["TE_SB_dis"]] <- merge(tables[["TE_dis"]], tables[["SB_dis"]], by = c("year", "PMID"))
tables_mixed[["TE_Omic_dis"]] <- merge(tables[["TE_dis"]], tables[["Omic_gen"]], by = c("year", "PMID"))
# All
# Get intersection of the three
tables_mixed[["all_gen"]] <- merge(tables_mixed[["TE_SB_gen"]], tables_mixed[["TE_Omic_gen"]], 
                               by = c("PMID", "year", "category.x"))
colnames(tables_mixed[["all_gen"]]) <- c("PMID", "year", "TE", "SB", "Omic")
# Save object!
saveRDS(tables_mixed, file = paste0(dir_out, "query_tables_mixed.rds"), compress = TRUE, refhook = NULL)
```

## Intersect PMC categories

PMC search return way more results since it's querying the full text if possible.
A lot of results from math modelling have nothing to do with bioengineering, meaning
the search is broader but less specific. So we only focus on intersections with the 
tissue eng categories and look those up.

```{r}
# Intersect general queries (not specific to disease)
tables_mixed[["TE_SB_gen_PMC"]] <- merge(tables[["TE_gen_PMC"]], tables[["SB_gen_PMC"]], by = c("year", "PMC"))
tables_mixed[["TE_Omic_gen_PMC"]] <- merge(tables[["TE_gen_PMC"]], tables[["Omic_gen_PMC"]], by = c("year", "PMC"))
# Intersect disease queries
tables_mixed[["TE_SB_dis_PMC"]] <- merge(tables[["TE_dis_PMC"]], tables[["SB_dis_PMC"]], by = c("year", "PMC"))
tables_mixed[["TE_Omic_dis_PMC"]] <- merge(tables[["TE_dis_PMC"]], tables[["Omic_gen_PMC"]], by = c("year", "PMC"))
# Save object!
saveRDS(tables_mixed, file = paste0(dir_out, "query_tables_mixed.rds"), compress = TRUE, refhook = NULL)
```


## Translate PMC to PMID and get rid of reviews

We retrieved a lot of random PMCs so this step is slow. We only translate papers
that intersect with TE since those are the interesting ones. We can save the dictionary 
and read it every time so that we don't re-make it if we change queries, but we 
just update it with new PMC-PMID. The dictionary is saved during each batch

```{r}
# Load tables
tables_mixed <- readRDS(paste0(dir_out, "query_tables_mixed.rds"))
# Get PMC - PMID dictionary
dictionary <- get_PMID_PMC_batched(c(tables_mixed[["TE_SB_gen_PMC"]]$PMC,
                                   tables_mixed[["TE_Omic_gen_PMC"]]$PMC,
                                   tables[["PMC_extra"]]$PMC) %>% 
                                   unique(),
                                   do.clean = T,
                                   save_name = "../Data/dictionary.txt")

# Or load directly
dictionary <- read.table("../Data/dictionary.txt", header = T, sep = "\t")

# Apply dictionary to intersections - this adds a PMID and gets rid of reviews and other junk
tables_mixed[["TE_SB_gen_PMC"]] <- merge(tables_mixed[["TE_SB_gen_PMC"]], dictionary, by = "PMC")
tables_mixed[["TE_Omic_gen_PMC"]] <- merge(tables_mixed[["TE_Omic_gen_PMC"]], dictionary, by = "PMC")
tables_mixed[["TE_SB_dis_PMC"]] <- merge(tables_mixed[["TE_SB_dis_PMC"]], dictionary, by = "PMC")
tables_mixed[["TE_Omic_dis_PMC"]] <- merge(tables_mixed[["TE_Omic_dis_PMC"]], dictionary, by = "PMC")

# Get intersection of the three
tables_mixed[["all_gen_PMC"]] <- merge(tables_mixed[["TE_SB_gen_PMC"]],
                                       tables_mixed[["TE_Omic_gen_PMC"]], 
                                       by = c("PMC", "PMID", "year", "category.x"))
colnames(tables_mixed[["all_gen_PMC"]]) <- c("PMC", "PMID", "year", "TE", "SB", "Omic")
# Translate extra PMC queries and store
tables[["PMC_extra"]] <- merge(tables[["PMC_extra"]], dictionary, by = "PMC")
# Save tables
saveRDS(tables_mixed, file = paste0(dir_out, "query_tables_mixed.rds"), compress = TRUE, refhook = NULL)
```

# Verify queries

Manually verify that the titles of the full intersection are relevant examples.
Get a list of unique PMID/PMC and get their titles and prompt user for manual 
confirmation. This part MUST be interactive, so run separately if needed. We included 
it to make it easier to go through multiple titles of papers for QC purposes.

We can re-read data here to skip all of the chunks above

```{r}
tables_mixed <- readRDS(file = paste0(dir_out, "query_tables_mixed.rds"))
```

Also, some review papers are not indexed as "review" in pubmed so they don't get
filtered out. As a flag, we can check if the papers have the term "review" in the 
abstract.

```{r}
# Do in chunks of 50 (although internally stuff is saved every 10 prompts)
PMC_all_gen_unique <- verify_titles(tables_mixed[["all_gen_PMC"]]$PMID %>% unique(), 
                      file_prechecked =  paste0(dir_out,"checkedID/PMC_all_unique_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/PMC_all_unique_checked.txt"),
                      batch_size = 50)

```

Now check pubmed queries. There are fewer of these. 

```{r}
# Do in chunks of 50 (although internally stuff is saved every 10 prompts)
PMID_all_gen_unique <- verify_titles(tables_mixed[["all_gen"]]$PMID %>% unique(), 
                      file_prechecked =  paste0(dir_out,"checkedID/PMID_all_unique_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/PMID_all_unique_checked.txt"),
                      batch_size = 50)

# Do for intersections. No need to return stuff
verify_titles(tables_mixed[["TE_SB_gen"]]$PMID %>% unique(), 
                      file_prechecked =  paste0(dir_out,"checkedID/PMID_TE_SB_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/PMID_TE_SB_checked.txt"),
                      batch_size = 50)

```

Concatenate the selected PMIDs and export to text file. Then, paste to excel manually and 
double-check papers for relevance

```{r}
PMC_all_gen_unique$dataset <- "PMC"
PMID_all_gen_unique$dataset <- "PMID"

rbind(filter(PMID_all_gen_unique, is.good == "y" & is.review == FALSE),
      filter(PMC_all_gen_unique, is.good == "y" & is.review == FALSE)) %>%
  write.table(paste0(dir_out,'PMID_to_check.txt'), col.names = F, row.names = F, quote = F, sep = "\t")
```

# Analyze and clean results

## Get overlaps for PMID

Clean up the categories for math models a bit. Lump together diff eqs, dynamic models
and PK models since they are all based on differential equations. For the category "mathematical models",
exclude all IDs that are in another category to reduce overlaps since this category is non-specific. 
Lump hydrogels into engineered tissue models. This could be done from the start by combining queries but the queries are already too long!

```{r}
# Making a tiny function to avoid repeating code
clean_SB_table <- function(table){
  old_ID <- colnames(table)[3]
  colnames(table)[3] <- "ID"
  # Lump together ODEs, dynamic models, PK models, as diff eqs
  table$category[table$category %in% 
                   c("Differential equation", "Dynamic model", "PK")] <- "Differential equation"
  # Remove items from "mathematical models" that are in another category already
  table2 <- table %>% filter(category != "Mathematical models")
  table3 <- table %>% filter(category == "Mathematical models")
  
  table <- rbind(table2,
                 table3 %>% filter(!(ID %in% table2$ID))
                 )
  
  colnames(table)[3] <- old_ID
  return(table)
}

clean_TE_table <- function(table){
  old_ID <- colnames(table)[3]
  colnames(table)[3] <- "ID"
  # Lump together ODEs, dynamic models, PK models, as diff eqs
  table$category[table$category == "Hydrogels"] <- "3D Engineered model"
  colnames(table)[3] <- old_ID
  return(table)
}

```

Producing a list of plots. These wrapper functions save the data structures that render the plot,
but the actual plots are rendered in the following chunk

```{r}
# Get overlaps between categories
overlaps_dis <- get_queries_overlap_lists(list(TE = clean_TE_table(tables$TE_dis),
                                              MM = clean_SB_table(tables$SB_dis),
                                              Omics = tables$Omic_dis
                                              ),
                                          save_name = paste0(dir_out, "/Plots/overlaps_dis.rds"))

overlaps_gen <- get_queries_overlap_lists(list(TE = clean_TE_table(tables$TE_gen),
                                              MM = clean_SB_table(tables$SB_gen),
                                              Omics = tables$Omic_gen
                                              ),
                                          save_name = paste0(dir_out, "/Plots/overlaps_gen.rds"))

overlaps_dis_PMC <- get_queries_overlap_lists(list(TE = clean_TE_table(tables$TE_dis_PMC),
                                              MM = clean_SB_table(tables$SB_dis_PMC),
                                              Omics = tables$Omic_dis_PMC
                                              ),
                                          save_name = paste0(dir_out, "/Plots/overlaps_dis_PMC.rds"))
```

Read from file instead

```{r}

# Read plot files 
overlaps_gen <- readRDS(paste0(dir_out, "/Plots/overlaps_gen.rds"))
overlaps_dis <- readRDS(paste0(dir_out, "/Plots/overlaps_dis.rds"))
overlaps_dis_PMC <- readRDS(paste0(dir_out, "/Plots/overlaps_dis_PMC.rds"))

```

Created a small wrapper function that specifies plot size based on what will fit a page nicely

```{r}

# Plot for the different reference queries that we overlapped
# Wrapper function to avoid repeating stuff
make_plots <- function(overlap_table, dir_out){

ggsave(paste0(dir_out,"Omic_overlap.pdf"), 
       plot = overlap_table[["Omics_Omics"]][["plot"]] +
                  theme(text = element_text(size = size_text),
                        axis.text.y = element_text(color = "black"),
                        axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
                        legend.key.size = unit(size_legend, "cm")) ,
      units = "cm", width = 12, height = 8)

ggsave(paste0(dir_out,"TE_overlap.pdf"), 
       plot = overlap_table[["TE_TE"]][["plot"]] +
                  theme(text = element_text(size = size_text),
                        axis.text.y = element_text(color = "black"),
                        axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
                        legend.key.size = unit(size_legend, "cm")) ,
      units = "cm", width = 8, height = 8)

ggsave(paste0(dir_out,"SB_overlap.pdf"), 
       plot = overlap_table[["MM_MM"]][["plot"]] +
                  theme(text = element_text(size = size_text),
                        axis.text.y = element_text(color = "black"),
                        axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
                        legend.key.size = unit(size_legend, "cm")) ,
      units = "cm", width = 14, height = 10)



# Reorder axes and save - lots of repetitive code that I could condense later

ggsave(paste0(dir_out,"TE_Omic_overlap.pdf"), 
       plot = mod_axes(overlap_table[["TE_Omics"]][["plot"]], 
                       my_query_TE$category[my_query_TE$category != "Hydrogels"],
                       my_query_Omic$category), 
       units = "cm", width = 8, height = 10)

ggsave(paste0(dir_out,"TE_SB_overlap.pdf"), 
       plot = mod_axes(overlap_table[["TE_MM"]][["plot"]],
                       my_query_TE$category[my_query_TE$category != "Hydrogels"],
                       my_query_SB$category[!(my_query_SB$category %in% c("PK", "Dynamic model"))]), 
       units = "cm", width = 8, height = 10)

ggsave(paste0(dir_out,"Omic_SB_overlap.pdf"), 
       plot = mod_axes(overlap_table[["MM_Omics"]][["plot"]],
                       my_query_SB$category[!(my_query_SB$category %in% c("PK", "Dynamic model"))],
                       my_query_Omic$category), 
       units = "cm", width = 12, height = 10)

# Venn diagram intersecting main categories
ggsave(paste0(dir_out,"Venn.pdf"), 
       plot = overlap_table[["Venn"]], 
       units = "cm", width = 7, height = 7)

# Make additional plot for supp info: Dotplots summarizing the Jaccard index of the queries
dummy <- rbind(data.frame(category = "TE", 
                          overlap_table[["TE_TE"]][["table"]]),
               data.frame(category = "MM", 
                          overlap_table[["MM_MM"]][["table"]]),
               data.frame(category = "Omics", 
                          overlap_table[["Omics_Omics"]][["table"]]))
# Remove duplicated pairs (A B = B A) and plot
dummy$pairs <- apply(dummy, MARGIN = 1, 
                     FUN = function(x){str_sort(c(x[2], x[3])) %>% paste(collapse = " ")})

plt_histo <- dummy %>% 
  filter(jaccard < 0.9999 & duplicated(pairs) & count_both > 0) %>% 
  ggplot(aes(x = factor(category), y = 100*jaccard)) + 
  geom_boxplot(outlier.alpha = 0, color = "black", size = size_line) +  
  geom_jitter(width = 0.1, size = size_dot, color = "steelblue", alpha = 0.7) +
  labs(x = NULL, y = "Jaccard indices (%)") + 
  scale_x_discrete(labels = c("Math models", "Exp methods", "3D in vitro"))

ggsave(paste0(dir_out,"Jaccard_dot.pdf"), 
       plot = plt_histo, 
       units = "cm", width = 7.5, height = 7)

}

# Render plots and export pdf
make_plots(overlaps_dis, paste0(dir_out,"Plots/PMID_dis/"))
make_plots(overlaps_gen,paste0(dir_out,"Plots/PMID_gen/"))
make_plots(overlaps_dis_PMC, paste0(dir_out,"Plots/PMC_dis/"))
```
