---
title: "Plots and analysis of Pubmed queries"
author: "Jose L. Cadavid"
output: html_notebook
---

This notebook loads tables of pubmed IDs that come from multiple queries performed
in the "run_pubmed_queries" notebook. Most of the chunks can be run fully automatically,
except the last few where user input is needed when looking at article titles and abstracts.


# Load packages

```{r}
library(readxl)
library(ggvenn)
library(xlsx)
library(scales)
```

# Load data

Define directory where files are contained and run extra functions for plotting and so on.

Note: Query tables last updated April 26, 2023

```{r}
# Load extra functions - functions for plotting or wrapper functions
source("extra_functions.R")
# Set Entrez key
set_entrez_key(api_key)
# Define directories - create output directory by hand before running this
query_file <- "queries.xlsx"
dir_out <- "Results2023/"
tables <- readRDS(file = paste0(dir_out, "query_tables.rds"))
```

## Intersect categories

For all unique PMID and PMC retrieved from the initial queries, see if they are also in the 
tissue eng, systems bio, or cell-cell categories

THIS PART IS NOT NECESSARY. MODIFY

```{r}

# Get unique PMIDs retrieved (ALL of them) and whether they are in the SB or TE categories
PMID_unique <- c(tables$SB_gen$PMID, 
                     tables$TE_gen$PMID, 
                     tables$Omic_gen$PMID) %>% unique()

PMID_unique <- data.frame(PMID = PMID_unique, 
                          is.sys_bio = PMID_unique %in% PMID_systems_bio, 
                          is.tissue_eng = PMID_unique %in% PMID_tissue_eng,
                          is.cell_cell = PMID_unique %in% PMID_cell_cell)
# Now PMC
PMC_unique <- c(tables$SB_gen_PMC$PMC, 
                     tables$TE_gen_PMC$PMC, 
                     tables$Omic_gen_PMC$PMC) %>% unique()

PMC_unique <- data.frame(PMC = PMC_unique, 
                          is.sys_bio = PMC_unique %in% PMC_systems_bio, 
                          is.tissue_eng = PMC_unique %in% PMC_tissue_eng,
                          is.cell_cell = PMC_unique %in% PMC_cell_cell)

```

## Intersect PMID categories

Intersect interesting categories to make plots later on or to check interesting IDs manually

```{r}
tables_mixed <- NULL
tables_mixed[["TE_SB_gen"]] <- merge(tables[["TE_gen"]], tables[["SB_gen"]], by = c("year", "PMID"))
tables_mixed[["TE_Omic_gen"]] <- merge(tables[["TE_gen"]], tables[["Omic_gen"]], by = c("year", "PMID"))
# Intersect disease queries
tables_mixed[["TE_SB_dis"]] <- merge(tables[["TE_dis"]], tables[["SB_dis"]], by = c("year", "PMID"))
tables_mixed[["TE_Omic_dis"]] <- merge(tables[["TE_dis"]], tables[["Omic_gen"]], by = c("year", "PMID"))
# All
# Get intersection of the three
tables_mixed[["all_gen"]] <- merge(tables_mixed[["TE_SB_gen"]], tables_mixed[["TE_Omic_gen"]], 
                               by = c("PMID", "year", "category.x"))
colnames(tables_mixed[["all_gen"]]) <- c("PMID", "year", "TE", "SB", "Omic")
# Save object!
saveRDS(tables_mixed, file = paste0(dir_out, "query_tables_mixed.rds"), compress = TRUE, refhook = NULL)
```

## Intersect PMC categories

PMC search return way more results since it's querying the full text if possible.
A lot of results from math modelling have nothing to do with bioengineering, meaning
the search is broader but less specific. So we only focus on intersections with the 
tissue eng categories and look those up.

```{r}
# Intersect general queries (not specific to disease)
tables_mixed[["TE_SB_gen_PMC"]] <- merge(tables[["TE_gen_PMC"]], tables[["SB_gen_PMC"]], by = c("year", "PMC"))
tables_mixed[["TE_Omic_gen_PMC"]] <- merge(tables[["TE_gen_PMC"]], tables[["Omic_gen_PMC"]], by = c("year", "PMC"))
# Intersect disease queries
tables_mixed[["TE_SB_dis_PMC"]] <- merge(tables[["TE_dis_PMC"]], tables[["SB_dis_PMC"]], by = c("year", "PMC"))
tables_mixed[["TE_Omic_dis_PMC"]] <- merge(tables[["TE_dis_PMC"]], tables[["Omic_gen_PMC"]], by = c("year", "PMC"))
# Save object!
saveRDS(tables_mixed, file = paste0(dir_out, "query_tables_mixed.rds"), compress = TRUE, refhook = NULL)
```


## Translate PMC to PMID and get rid of reviews

We retrieved a lot of random PMCs so this step is slow. We only translate papers
that intersect with TE since those are the interesting ones. We can save the dictionary 
and read it every time so that we don't re-make it if we change queries, but we 
just update it with new PMC-PMID. The dictionary is saved during each batch

```{r}
# Create dictionary
dictionary <- get_PMID_PMC_batched(c(tables_mixed[["TE_SB_gen_PMC"]]$PMC,
                                     tables_mixed[["TE_Omic_gen_PMC"]]$PMC,
                                     tables[["PMC_extra"]]$PMC) %>% 
                                     unique(),
                                   do.clean = T,
                                   save_name = paste0(dir_out,"dictionary.txt"))

dictionary <- read.table(paste0(dir_out,"dictionary.txt"), header = T)


# Apply dictionary to intersections - this adds a PMID and gets rid of reviews and other junk
tables_mixed[["TE_SB_gen_PMC"]] <- merge(tables_mixed[["TE_SB_gen_PMC"]], dictionary, by = "PMC")
tables_mixed[["TE_Omic_gen_PMC"]] <- merge(tables_mixed[["TE_Omic_gen_PMC"]], dictionary, by = "PMC")
tables_mixed[["TE_SB_dis_PMC"]] <- merge(tables_mixed[["TE_SB_dis_PMC"]], dictionary, by = "PMC")
tables_mixed[["TE_Omic_dis_PMC"]] <- merge(tables_mixed[["TE_Omic_dis_PMC"]], dictionary, by = "PMC")

# Get intersection of the three
tables_mixed[["all_gen_PMC"]] <- merge(tables_mixed[["TE_SB_gen_PMC"]], tables_mixed[["TE_Omic_gen_PMC"]], 
                                 by = c("PMC", "PMID", "year", "category.x"))
colnames(tables_mixed[["all_gen_PMC"]]) <- c("PMC", "PMID", "year", "TE", "SB", "Omic")
# Translate extra PMC queries and store
tables[["PMC_extra"]] <- merge(tables[["PMC_extra"]], dictionary, by = "PMC")
# Save tables
saveRDS(tables_mixed, file = paste0(dir_out, "query_tables_mixed.rds"), compress = TRUE, refhook = NULL)

```

# Verify queries

Manually verify that the titles of the full intersection are relevant examples.
Get a list of unique PMID/PMC and get their titles and prompt user for manual 
confirmation. This part MUST be interactive, so run separately if needed. We included 
it to make it easier to go through multiple titles of papers for QC purposes.

We can re-read data here to skip all of the chunks above

```{r}
tables_mixed <- readRDS(file = paste0(dir_out, "query_tables_mixed.rds"))
```


Also, some review papers are not indexed as "review" in pubmed so they don't get
filtered out. As a flag, we can check if the papers have the term "review" in the 
abstract.


```{r}
# Do in chunks of 50 (although internally stuff is saved every 10 prompts)
PMC_all_gen_unique <- verify_titles(tables_mixed[["all_gen_PMC"]]$PMID %>% unique(), 
                      file_prechecked =  paste0(dir_out,"checkedID/PMC_all_unique_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/PMC_all_unique_checked.txt"),
                      batch_size = 50)

```

Now check pubmed queries. There are fewer of these. 

```{r}
# Do in chunks of 50 (although internally stuff is saved every 10 prompts)
PMID_all_gen_unique <- verify_titles(tables_mixed[["all_gen"]]$PMID %>% unique(), 
                      file_prechecked =  paste0(dir_out,"checkedID/PMID_all_unique_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/PMID_all_unique_checked.txt"),
                      batch_size = 50)

# Do for intersections. No need to return stuff
verify_titles(tables_mixed[["TE_SB_gen"]]$PMID %>% unique(), 
                      file_prechecked =  paste0(dir_out,"checkedID/PMID_TE_SB_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/PMID_TE_SB_checked.txt"),
                      batch_size = 50)

```

Now check IDs retrieved from intersecting systems-bio and tissue-eng mesh headers.
Some of these might have been checked in the previous two chunks, so we load those
pre-checked files. PMC need to be translated to PMID, but we already have a dictionary 
for those that will exclude reviews and other junk.

```{r}
# Do in chunks of 50 (although internally stuff is saved every 10 prompts)
extra_all_gen_unique <- verify_titles(tables[["extra_IDs"]]$PMID %>% unique(),
                      file_prechecked =  paste0(dir_out,"checkedID/extra_all_unique_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/extra_all_unique_checked.txt"),
                      batch_size = 50)

```
Now check IDs that are related to co-cultures or cell-cell comm. Not lumping
with the other extra_IDs since co-culture or cell-cell is using different categories
and intersecting with the tisse eng and SB categories. Not looking at the intersection
between TE, Omic and SB, rather intersections of co-cultures and cell-cell to understand
the extent at which co-cultures are discussed. Only for PMID since we already have a long list 
of references to go through

NOT UPDATED

```{r}

coculture_all_gen_unique <- verify_titles(tables$PMID_co[tables$PMID_co %in% 
                                                           union(tables$TE_SB_gen$PMID, tables$TE_Omic_gen$PMID)] ,
                      file_prechecked =  paste0(dir_out,"checkedID/coculture_unique_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/coculture_unique_checked.txt"),
                      batch_size = 50)

cellcell_all_gen_unique <- verify_titles(tables$PMID_cell_cell[tables$PMID_cell_cell %in% 
                                                           union(tables$TE_SB_gen$PMID, tables$TE_Omic_gen$PMID)] ,
                      file_prechecked =  paste0(dir_out,"checkedID/cellcell_unique_checked.txt"),
                      file_store = paste0(dir_out,"checkedID/cellcell_unique_checked.txt"),
                      batch_size = 50)


```


Concatenate the selected PMIDs and export to text file. Then, paste to excel manually and 
double-check papers for relevance

```{r}
PMC_all_gen_unique$dataset <- "PMC"
PMID_all_gen_unique$dataset <- "PMID"
extra_all_gen_unique$dataset <- "extra"

rbind(filter(PMID_all_gen_unique, is.good == "y" & is.review == FALSE),
      filter(PMC_all_gen_unique, is.good == "y" & is.review == FALSE),
      filter(extra_all_gen_unique, is.good == "y" & is.review == FALSE)) %>%
  write.table(paste0(dir_out,'PMID_to_check.txt'), col.names = F, row.names = F, quote = F, sep = "\t")


```



# Analyze and clean results

## Get trends

Using the info from the years to get counts of papers per category. We eliminate duplicates
within each category to keep the first publication date (some papers show up twice), then summarize 
counts and plot. Using a wrapper function.

UNFINISHED

```{r}
# Initialize list for results
results <- NULL

# Plot trends separated by query category
results[["trend_TE_gen"]] <- plot_trends(tables$TE_gen, var_group2 = "category")
results[["trend_SB_gen"]] <- plot_trends(tables$SB_gen, var_group2 = "category")
results[["trend_Omic_gen"]] <- plot_trends(tables$Omic_gen, var_group2 = "category")

# Trends of overlaps
results[["trend_TE_SB_gen"]] <- plot_trends(tables$TE_SB_gen)
results[["trend_TE_Omic_gen"]] <- plot_trends(tables$TE_Omic_gen)



```


## Get overlaps for PMID

Clean up the categories for math models a bit. Lump together diff eqs, dynamic models
and PK models since they are all based on differential equations. For the category "mathematical models",
exclude all IDs that are in another category to reduce overlaps since this category is non-specific.

Lump hydrogels into engineered tissue models

This could be done from the start by combining queries but the queries are already too long!

Making a tiny function to avoid repeating code

```{r}
clean_SB_table <- function(table){
  old_ID <- colnames(table)[3]
  colnames(table)[3] <- "ID"
  # Lump together ODEs, dynamic models, PK models, as diff eqs
  table$category[table$category %in% 
                   c("Differential equation", "Dynamic model", "PK")] <- "Differential equation"
  # Remove items from "mathematical models" that are in another category already
  table2 <- table %>% filter(category != "Mathematical models")
  table3 <- table %>% filter(category == "Mathematical models")
  
  table <- rbind(table2,
                 table3 %>% filter(!(ID %in% table2$ID))
                 )
  
  colnames(table)[3] <- old_ID
  return(table)
}

clean_TE_table <- function(table){
  old_ID <- colnames(table)[3]
  colnames(table)[3] <- "ID"
  # Lump together ODEs, dynamic models, PK models, as diff eqs
  table$category[table$category == "Hydrogels"] <- "3D Engineered model"
  colnames(table)[3] <- old_ID
  return(table)
}

```



Producing a list of plots. The following chunks actually save the plots

```{r}
# Get overlaps between categories
overlaps_dis <- get_queries_overlap_lists(list(TE = clean_TE_table(tables$TE_dis),
                                              MM = clean_SB_table(tables$SB_dis),
                                              Omics = tables$Omic_dis
                                              ),
                                          save_name = paste0(dir_out, "/Plots/overlaps_dis.rds"))

overlaps_gen <- get_queries_overlap_lists(list(TE = clean_TE_table(tables$TE_gen),
                                              MM = clean_SB_table(tables$SB_gen),
                                              Omics = tables$Omic_gen
                                              ),
                                          save_name = paste0(dir_out, "/Plots/overlaps_gen.rds"))

overlaps_dis_PMC <- get_queries_overlap_lists(list(TE = clean_TE_table(tables$TE_dis_PMC),
                                              MM = clean_SB_table(tables$SB_dis_PMC),
                                              Omics = tables$Omic_dis_PMC
                                              ),
                                          save_name = paste0(dir_out, "/Plots/overlaps_dis_PMC.rds"))
```

Read from file instead
```{r}

# Read plot files 
overlaps_gen <- readRDS(paste0(dir_out, "/Plots/overlaps_gen.rds"))
overlaps_dis <- readRDS(paste0(dir_out, "/Plots/overlaps_dis.rds"))
overlaps_dis_PMC <- readRDS(paste0(dir_out, "/Plots/overlaps_dis_PMC.rds"))

```

```{r}

# Plot for the different reference queries that we overlapped


# Wrapper function to avoid repeating stuff
make_plots <- function(overlap_table, dir_out){

ggsave(paste0(dir_out,"Omic_overlap.pdf"), 
       plot = overlap_table[["Omics_Omics"]][["plot"]] +
                  theme(text = element_text(size = size_text),
                        axis.text.y = element_text(color = "black"),
                        axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
                        legend.key.size = unit(size_legend, "cm")) ,
      units = "cm", width = 12, height = 8)

ggsave(paste0(dir_out,"TE_overlap.pdf"), 
       plot = overlap_table[["TE_TE"]][["plot"]] +
                  theme(text = element_text(size = size_text),
                        axis.text.y = element_text(color = "black"),
                        axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
                        legend.key.size = unit(size_legend, "cm")) ,
      units = "cm", width = 8, height = 8)

ggsave(paste0(dir_out,"SB_overlap.pdf"), 
       plot = overlap_table[["MM_MM"]][["plot"]] +
                  theme(text = element_text(size = size_text),
                        axis.text.y = element_text(color = "black"),
                        axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
                        legend.key.size = unit(size_legend, "cm")) ,
      units = "cm", width = 14, height = 10)



# Reorder axes and save - lots of repetitive code that I could condense later
ggsave(paste0(dir_out,"TE_Omic_overlap.pdf"), 
       plot = mod_axes(overlap_table[["TE_Omics"]][["plot"]], 
                       my_query_TE$category[my_query_TE$category != "Hydrogels"],
                       my_query_Omic$category), 
       units = "cm", width = 8, height = 10)

ggsave(paste0(dir_out,"TE_SB_overlap.pdf"), 
       plot = mod_axes(overlap_table[["TE_MM"]][["plot"]],
                       my_query_TE$category[my_query_TE$category != "Hydrogels"],
                       my_query_SB$category[!(my_query_SB$category %in% c("PK", "Dynamic model"))]), 
       units = "cm", width = 8, height = 10)

ggsave(paste0(dir_out,"Omic_SB_overlap.pdf"), 
       plot = mod_axes(overlap_table[["MM_Omics"]][["plot"]],
                       my_query_SB$category[!(my_query_SB$category %in% c("PK", "Dynamic model"))],
                       my_query_Omic$category), 
       units = "cm", width = 12, height = 10)


ggsave(paste0(dir_out,"Venn.pdf"), 
       plot = overlap_table[["Venn"]], 
       units = "cm", width = 7, height = 7)
}

# Make the plots
make_plots(overlaps_dis, paste0(dir_out, "/Plots/PMID_dis/"))
make_plots(overlaps_gen, paste0(dir_out, "/Plots/PMID_gen/"))
make_plots(overlaps_dis_PMC, paste0(dir_out, "/Plots/PMC_dis/"))



```






# Get abstracts of papers in the intersections with tissue eng

We don't care as much about reviewing math modelling with Omic data

```{r}
ID_for_abstracts <- rbind(tables$TE_SB_dis,
                          tables$TE_Omic_dis) %>% 
                    group_by(PMID) %>%
                    summarize(year = min(year))

PMID_dis_abstracts <- get_title_abstract(ID_for_abstracts$PMID,
                                save_name = paste0(dir_out,"PMID_title_abstract.txt"))
# Add year info
PMID_dis_abstracts <- merge(PMID_dis_abstracts, ID_for_abstracts, by = "PMID")
write.table(PMID_dis_abstracts, paste0(dir_out,"PMID_title_abstract.txt"), quote = T, row.names = F, sep = "\t")
```


# For thesis - additional TE queries focused on PDAC or cancer only

```{r}

add_query_mod <- " AND (pancreatic[tiab] OR pancreas[tiab]) AND (adenocarcinoma[tiab] OR cancer[tiab] OR carcinoma[tiab]) NOT (review[pt] OR comment[pt] OR editorial[pt] OR systematic-review[pt] OR comparative-study[pt]) AND english[lang]"

table_TE_PDAC <- get_queries_range(my_query_TE, add_query_mod, years, "table_TE_PDAC", dir_out = dir_out)

add_query_mod <- " AND (cancer[tiab] OR carcinoma[tiab]) NOT (review[pt] OR comment[pt] OR editorial[pt] OR systematic-review[pt] OR comparative-study[pt]) AND english[lang]"

table_TE_cancer <- get_queries_range(my_query_TE, add_query_mod, years, "table_TE_cancer", dir_out = dir_out)

# Summarize counts in each category
table_TE_merge <- merge(table_TE_PDAC %>% group_by(category) %>% summarize(PDAC = sum(!is.na(PMID))),
                        table_TE_cancer %>% group_by(category) %>% summarize(Cancer = sum(!is.na(PMID))),
                        by = "category")

# Plot total counts
plt_TE_PDAC <- table_TE_merge %>% 
                pivot_longer(cols = c("PDAC", "Cancer"), names_to = "type", values_to = "count") %>%
                filter(count > 0) %>%
                ggplot(aes(x = as.factor(category), y = count, fill = as.factor(type))) +
                  geom_col(position = "dodge2") +
                  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                                labels = trans_format("log10", math_format(10^.x)),
                                limits = c(1,10^4)) +
                  labs(x = NULL, y = "PubMed articles", fill = NULL) +
                  scale_fill_manual(values = c("gray", "steelblue")) +
                  scale_x_discrete(limits = my_query_TE$category)

plt_TE_PDAC <- add_theme(plt_TE_PDAC) #+ theme(axis.text.x = element_text(color = "black", angle = 0, hjust = 0))
# Save
ggsave(paste0(dir_out,"plot_TE_PDAC.pdf"), plot = plt_TE_PDAC, units = "cm", width = 19, height = 4)

# Get overlap in categories
TE_overlap_cancer <- get_queries_overlap(table_TE_cancer)
plt_TE_overlap <- TE_overlap_cancer[["plot"]] +
                  scale_y_discrete(limits = my_query_TE$category) +
                  scale_x_discrete(limits = my_query_TE$category) +
                  labs(size = "Jaccard index (%)")

# Override theme
plt_TE_overlap <- plt_TE_overlap +
                  theme(text = element_text(size = size_text),
                        axis.text.y = element_text(color = "black"),
                        axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
                        legend.key.size = unit(size_legend, "cm")) 

ggsave(paste0(dir_out,"plot_TE_overlap_thesis.pdf"), plot = plt_TE_overlap, units = "cm", width = 10, height = 8)

```

