---
title: "Trend analysis of systems biology"
author: "Jose L. Cadavid"
output: html_notebook
---

This notebook looks at pubmed abstracts that have the word "systems" in them and
then to find names of fields related to systems biology

# Load libraries

```{r}
library(tidytext, quietly = T)
library(forcats, quietly = T)
library(ggrepel, quietly = T)
library(cowplot, quietly = T)
library(scales, quietly = T)
data(stop_words)
setwd(getwd())
source("extra_functions.R")
```

# Get words associated with "systems"

Get IDs of papers that mention the word "systems" in their title. Systems biology 
is first discussed around 2000 so look starting then. Only looking at reviews
to reduce the list under the assumption that relevant fields have been reviewed

```{r}
IDs <- NULL
for (year in seq(1999,2023)){
  term <- paste0('systems[title] AND review[pt] AND eng[lang] AND hasabstract')
  ID_new <- get_IDs_batched(query = term, year = year, data_base = "pubmed")
  IDs <- c(IDs, ID_new)
}
```

Get titles and abstracts and look for words next to "systems"

```{r}
# Look in batches of 100
titles_list <- NULL
for (ii in seq(1, length(IDs), 100)){
  search <- entrez_summary(db = "pubmed", id = IDs[ii: min(ii+99, length(IDs))])
  titles <- lapply(search, FUN = function(x) x$title %>% tolower()) %>% unlist() 
  titles_list <- c(titles_list, titles)
}
# Extract words immediately following "systems". Exclude punctuation
words <- titles_list %>% 
          str_replace_all("[[:punct:]]"," ") %>% 
          str_replace_all("  "," ") %>%
          str_extract_all("(?<=systems ).*?(?= )") %>% 
          unlist()
# Make into data frame and remove common stop words
words_df <- data.frame(word = words) %>% 
            filter(!(word %in% stop_words$word)) %>%
            filter(!(word %in% c(" ", ""))) %>%
            group_by(word) %>% summarize(n = n())

words_df$word <- words_df$word %>% str_replace_all(" ","")

# Check whether pairs (systems + word) are indeed in the titles or if they were only
# included after removing a punctuation sign
words_df$is.title <- sapply(words_df$word, 
                            FUN = function(x) str_count(titles_list, paste("systems",x)) %>% sum())
```

After manual inspection we defined the following list of interesting terms. Now
we count the number of papers that have the combination of systems + word in title or abstract.
This part is a bit slow, so the results can be loaded in the next chunk if they have been created before.

```{r}
systems_terms <- data.frame(term = c("biology", "pharmacology", "medicine", "neuroscience", "genetics",
                 "metabolic engineering", "immunology", "toxicology", "vaccinology",
                 "chemistry", "physiology", "pathology", "biotechnology", "microbiology",
                 "biomedicine", "biochemistry", "glycobiology", "epidemiology", "genomics",
                 "oncology", "serology", "neurobiology", "biocatalysis", "virology"))

table <- NULL

# Get article counts for each term
for (ii in seq(1, nrow(systems_terms))){
  query <- paste0('"systems ', systems_terms$term[ii], '"[tiab] AND eng[la]')
  query_PMC <- paste0('"systems ', systems_terms$term[ii], '"[article]')
  for (year in seq(1999, 2023)){
    table <- rbind(table,
                   data.frame(term = systems_terms$term[ii],
                              year = year,
                              count = entrez_search(db = "pubmed", term = query, 
                                                    mindate = year, maxdate = year)$count,
                              count_PMC = entrez_search(db = "pmc", term = query_PMC, 
                                                        mindate = year, maxdate = year)$count))
  }
}
# Count sum of publications
table$term <- as.factor(table$term)
table <- table %>% group_by(term) %>% mutate(count_sum = cumsum(count), count_sum_PMC = cumsum(count_PMC)) %>% ungroup()

# Write table
write.table(table, file = "../Results/systems_words_table.txt", quote = F, row.names = F, sep = "\t")
```

Make nice plots

```{r}
# Read table of terms and counts
table <- read.table("../Results/systems_words_table.txt", header = T, sep = "\t")
# Plot
plt1 <- table %>% group_by(term) %>% filter(count_sum > 1) %>% ungroup() %>%
        ggplot(aes(x = year, y = count_sum)) + 
        geom_line(linewidth = 1.2, color = "steelblue") + 
        facet_wrap(facets = vars(term), ncol = 4) + 
        theme(legend.position = "none") + 
        scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x))) +
        scale_x_continuous(breaks = c(2000, 2010, 2020)) +
        labs(y = "Cumulative number of articles in Pubmed", x = "Year") +
        theme(text = element_text(size = size_text),
          axis.text.y = element_text(color = "black"),
          axis.text.x = element_text(color = "black"),
          legend.key.size = unit(size_legend, "cm"))



# Summary of table (year in which 5 total papers were crossed, and total number of papers)
table_sum <- table %>% group_by(term) %>% summarize(time_5 = year[count_sum > 5][1], 
                                                    total = max(count_sum), 
                                                    total_PMC = max(count_sum_PMC))

plt2 <- table_sum %>% mutate(term = fct_reorder(term, total)) %>%
        ggplot(aes(x = total, y = term)) + #, fill = log10(total)/(2022 - time_5))) +
        geom_col(fill = "steelblue") +
        scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x))) +
        scale_fill_viridis_c() + 
        labs(y = NULL, x = "Total count") +
        theme(text = element_text(size = size_text),
              axis.text.y = element_text(color = "black"),
              axis.text.x = element_text(color = "black"),
              legend.key.size = unit(size_legend, "cm"))

# Combine subfigures
plt_grid <- plot_grid(plt1, plt2, 
                      labels = c('a', 'b'), label_size = size_text, rel_widths = c(2, 1))

plt_grid
# Save
ggsave("../Results/Plots/systems_words_plot.pdf", plot = plt_grid, units = "cm", width = 17, height = 12)
ggsave("../Results/Plots/systems_words_plot.png", plot = plt_grid, units = "cm", width = 17, height = 12)
```
