---
title: "Scanning pubmed for relevant terms"
author: "Jose L. Cadavid"
output: html_notebook
---

This notebook opens an excel file with query tables and gets pubmed (and PMC) IDs
of relevant articles in a defined range of dates. It exports text files for the main tables,
and a consolidated RData file with all tables as a list. Not all queries and parameters are in the excel file,
but they can be seen in the first couple of chunks of this notebook (i.e. years, add_query)

It relies on the rentrez package and uses custom extra functions for looping and batching dates.
It also uses a custom api_key (each user should get their own) to allow for more queries in a given
time span. 

Unfortunately, sometimes the querying functions crash due to excessive calls to the entrez server, so
this notebook is meant to be run in a chunk-chunk or even line-line basis, as opposed to a single structured loop.

Running the whole thing can take around 1-2 hours. Results are meant to be analyzed in an accompanying notebook.

# Load packages


```{r}
library(readxl)
library(ggvenn)
library(xlsx)
library(scales)
```


# Define query file and output folder


```{r}
# Set directory to current path (opened by the notebook)
setwd(getwd())
# Load extra functions - functions for plotting or wrapper functions
source("extra_functions.R")
# Set Entrez key
set_entrez_key(api_key)
# Define directories - create output directory by hand before running this
query_file <- "queries.xlsx"
dir_out <- "Results2023/"
# Create subdirectories if they don't exist
sub_dirs <- c("Tables", "Plots", "checkedID")
for (ii in sub_dirs){
  if (!dir.exists(paste0(dir_out, ii))){
    dir.create(paste0(dir_out, ii))
  }
}
  

```


# Read queries

We perform a query for each term, even though in the end we will only care about 
intersections with tissue eng. This is done on purpose to get an idea of the magnitude of 
each field and to get trends in time, but we could simple get the intersections directly and save time.
We could also save time by not querying year-by-year, but then we wouldn't get
trend info which might come in handy.

We query with and without disease terms to get a broader idea of the field, but we
ultimately focus on disease. 

```{r}
# As a test one can put # in the queries to be ignored in the text file
# Get systems biology and modelling query terms
my_query_SB <- read_excel(query_file, sheet = "Modeling") %>% filter(!grepl("#", term))
# Get tissue engineering and in vitro models query terms
my_query_TE <- read_excel(query_file, sheet = "TissueEng") %>% filter(!grepl("#", term))
# Get omic and high throughput measurement terms
my_query_Omic <- read_excel(query_file, sheet = "Method") %>% filter(!grepl("#", term))

# Define query terms to specify disease states (the infix function %+% is a shorthand for paste0 that 
# concatenates a long multiline string (to make it easier to read the long query)
query_disease <- "disease OR diseases OR diseased OR " %+%
                  "pathology OR pathologies OR pathologic OR pathological OR " %+%
                  "pathophysiology OR pathophysiologies OR " %+% 
                  "cancer OR cancers OR carcinoma OR carcinomas OR neoplasm OR neoplasms"
# Common terms or restrictions to add to all queries. Exclude no abstract, reviews, etc
add_query_general <- "hasabstract AND eng[lang] NOT (review[pt] OR comment[pt] OR editorial[pt] " %+%
                      "OR systematic-review[pt] OR case-study[pt])"
# Add query terms for disease states
add_query_disease <- paste0('(', query_disease, ') AND (', add_query_general,')')

# PMC doesn't filter by publication type or language like pubmed, but it allows 
# excluding retracted papers from search. We enforce having an abstract to reduce trash
add_query_general_PMC <- '"has abstract"[filter] NOT "is retracted"[filter]'
add_query_disease_PMC <- paste0('(', query_disease, ') AND (', add_query_general_PMC, ')')

# Define year range to look for papers 
years <- 1990:2023

# Initialize list of tables: This will reduce the number of variables in the workspace
tables <- NULL

```

# Query PUBMED and PMC

## Get PUBMED query tables for each category

The function will internally save the full table, as well as a subsample table with 30 
papers from each category to be double-checked manually as a way of doing QC on the queries. 
Suffix ".txt" added internally to file name.
Note: We could put everything in a loop to make code concise by this way it's maybe more readable and each query can be checked independently at will (especially since the querying tends to crash once in a while)

```{r}
# Get general queries (not specifying cancer or disease)
tables[["SB_gen"]] <- get_queries_range(my_query_SB, add_query_general, years,
                                      data_base = "pubmed", "Tables/table_SB_general", 
                                      n_subsample = 30, dir_out = dir_out)
tables[["TE_gen"]] <- get_queries_range(my_query_TE, add_query_general, years, 
                                      data_base = "pubmed", "Tables/table_TE_general", 
                                      n_subsample = 30, dir_out = dir_out)
tables[["Omic_gen"]] <- get_queries_range(my_query_Omic, add_query_general, years, 
                                        data_base = "pubmed", "Tables/table_Omic_general", 
                                        n_subsample = 30, dir_out = dir_out)

# Get queries specific to disease by adding extra terms to query
tables[["SB_dis"]] <- get_queries_range(my_query_SB, add_query_disease, years, 
                                      data_base = "pubmed", "Tables/table_SB_disease", 
                                      n_subsample = 30, dir_out = dir_out)
tables[["TE_dis"]] <- get_queries_range(my_query_TE, add_query_disease, years, 
                                      data_base = "pubmed", "Tables/table_TE_disease", 
                                      n_subsample = 30, dir_out = dir_out)
tables[["Omic_dis"]] <- get_queries_range(my_query_Omic, add_query_disease, years, 
                                        data_base = "pubmed", "Tables/table_Omic_disease", 
                                        n_subsample = 30, dir_out = dir_out)


#tables[["Omic_dis"]] <- rio::import(paste0(dir_out, "/Tables/table_Omic_disease.txt"))

```


## Get PMC query tables for each category

PMC has some full manuscripts so the query can dig deeper into the paper. 
Ideally we would only parse the "methods" section, but the entrez utility doesn't 
do that. For now, we get a full query (remove [tiab] fields from query and change 
for [article]) and we clean afterwards. The [article] field does NOT query the abstracts
but abstracts are queryied by the PMID search.

```{r}
# Get general queries (not specifying cancer or disease)
tables[["SB_gen_PMC"]] <- get_queries_range(my_query_SB, add_query_general_PMC, years, 
                                          data_base = "pmc", "Tables/table_SB_general_PMC", 
                                          n_subsample = 30, dir_out = dir_out)
tables[["TE_gen_PMC"]] <- get_queries_range(my_query_TE, add_query_general_PMC, years, 
                                          data_base = "pmc","Tables/table_TE_general_PMC", 
                                          n_subsample = 30, dir_out = dir_out)
tables[["Omic_gen_PMC"]] <- get_queries_range(my_query_Omic, add_query_general_PMC, 
                                            years, data_base = "pmc","Tables/table_Omic_general_PMC", 
                                            n_subsample = 30, dir_out = dir_out)

# Get queries specific to disease by adding extra terms
tables[["SB_dis_PMC"]] <- get_queries_range(my_query_SB, add_query_disease_PMC, years, 
                                          data_base = "pmc","Tables/table_SB_disease_PMC", 
                                          n_subsample = 30, dir_out = dir_out)
tables[["TE_dis_PMC"]] <- get_queries_range(my_query_TE, add_query_disease_PMC, years, 
                                          data_base = "pmc","Tables/table_TE_disease_PMC", 
                                          n_subsample = 30, dir_out = dir_out)
tables[["Omic_dis_PMC"]] <- get_queries_range(my_query_Omic, add_query_disease_PMC, 
                                            years, data_base = "pmc","Tables/table_Omic_disease_PMC", 
                                            n_subsample = 30, dir_out = dir_out)

# Save object!
saveRDS(tables, file = paste0(dir_out, "query_tables.rds"), compress = TRUE, refhook = NULL)

```


## Check if PMIDs or PMCs are associated with other terms

This part is to see how many papers are indexed as tissue eng or systems bio, or
contain those keywords somewhere. The idea is to show how these fuzzy terms can miss
important work when just querying with them.

It's faster to just get a huge list of PMIDs of each category and see if they 
intersect with what we retrieved before, than querying each PMID and double-check it.

Also check for papers associated with cell-cell communication, and for queries 
associated with co-cultures (or multi-cultures)

```{r}
my_query_extra <- read_excel(query_file, sheet = "Extra") %>% filter(!grepl("#", term))


tables[["PMID_extra"]] <- get_queries_range(my_query_extra, 
                                            add_query_general, 
                                            years, 
                                            data_base = "pubmed")


tables[["PMC_extra"]] <- get_queries_range(my_query_extra %>% 
                                             mutate(term = str_replace_all(term, "tiab", "abstract")), 
                                          add_query_general_PMC, years, 
                                          data_base = "pmc")

# Save object!
saveRDS(tables, file = paste0(dir_out, "query_tables"), compress = TRUE, refhook = NULL)
```

Unused: queries for papers citing TCGA - Not fixed for updated functions

```{r}

# Query papers that cite TCGA as a resource
PMID_TCGA <- get_IDs_batched(query = paste0("(TCGA[tiab] OR cancer-genome-atlas[tiab])", 
                                           ' AND (', add_query_general, ') AND (', 
                                            years[1],':', tail(years,1), '[pdat])'), 
                                   data_base = "pubmed")

PMC_TCGA <- get_IDs_batched(query = paste0("(TCGA[article] OR cancer-genome-atlas[article])", 
                                           ' AND (', add_query_general_PMC, ') AND (', 
                                            years[1],':', tail(years,1), '[pdat])'), 
                                   data_base = "pmc")

# Store
tables[["PMID_TCGA"]] <- PMID_TCGA
tables[["PMC_TCGA"]] <- PMC_TCGA

# Check papers that have "disease model" and "in vitro" in pubmed abstracts
query_dis_model <- '((disease-model[tiab] OR disease-models[tiab] OR disease-modeling[tiab]) AND "in vitro"[tiab])'
PMID_disease_model <- get_IDs_batched(query = paste0(query_dis_model, 
                                           ' AND (', add_query_general, ') AND (', 
                                            years[1],':', tail(years,1), '[pdat])'), 
                                   data_base = "pubmed")

tables[["PMID_disease_model"]] <- PMID_disease_model


# Save object!
saveRDS(tables, file = paste0(dir_out, "query_tables"), compress = TRUE, refhook = NULL)

```

